---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->
# datasus

[![R-CMD-check](https://github.com/rpradosiqueira/datasus/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/rpradosiqueira/datasus/actions/workflows/R-CMD-check.yaml)

The "datasus" R package provides direct access to TABNET/DATASUS and
OpenDataSUS from R.
It covers vital statistics (SIM and SINASC), hospital production and
morbidity (SIH/SUS), ambulatory production (SIA/SUS), the National Registry
of Health Establishments (CNES), resident population estimates, and
notifiable conditions (SINAN). Historical immunization, nutritional
surveillance and financing tables and current SISCAN exam tables are also
available.

## Installation

The package is currently under active redevelopment and is not available from CRAN.
Install the development version from GitHub:

```{r , eval = FALSE}
install.packages("remotes")
remotes::install_github("rpradosiqueira/datasus")
```

## Example

SIM and SINASC now use the same catalog-driven interface as the other TABNET
systems:

```{r, eval = FALSE}
# Discover the mortality tables and inspect the current form
datasus_catalogo("sim")
datasus_opcoes("sim", "obitos", abrangencia = "municipio")

# Mortality by IBGE micro-region and ICD-10 chapter
sim(
  conjunto = "obitos",
  linha = "Microrregião IBGE",
  coluna = "Capítulo CID-10"
)

# State totals and live births within São Paulo municipalities
sim(abrangencia = "uf", periodo = 2024)
sinasc(uf = "SP", periodo = 2024)
```

The historical functions such as `sim_obt10_mun()` and `sinasc_nv_uf()` remain
as deprecated compatibility wrappers. New code should use `sim()` and
`sinasc()`.

The health-services API uses the same arguments for SIH, SIA and CNES:

```{r, eval = FALSE}
# Discover all supported datasets without accessing the network
datasus_catalogo()
datasus_catalogo("cnes")

# Inspect the live dimensions, measures, periods and filter values
op <- datasus_opcoes("sih", uf = "MS")
op$conteudo
op$filtros$carater_atendimento

# Hospital and ambulatory production
sih_producao(uf = "MS", conteudo = "Internações")
sia_producao(uf = "MS", conteudo = "Qtd.aprovada")

# Establishments and physical resources from CNES
cnes(uf = "MS")
cnes(conjunto = "leitos_internacao", uf = "MS")
```

Use `periodo = "last"` for the latest available competence, an exact label
such as `"Mai/2026"`, or a year such as `2025` to select all available
competences in that year. Named filters use the keys returned by
`datasus_opcoes()`.

Population, hospital morbidity and SINAN use the same query conventions:

```{r, eval = FALSE}
# Municipal population estimates; the official series currently ends in 2021
populacao_residente(uf = "MS", periodo = 2021)

# Hospital morbidity by ICD-10 chapter
sih_morbidade(
  uf = "MS",
  linha = "Capítulo CID-10",
  conteudo = "Internações",
  periodo = 2025
)

# Notifiable diseases and conditions
datasus_catalogo("sinan")
sinan("dengue", uf = "MS", periodo = 2025)
```

Additional health-service and surveillance tables:

```{r, eval = FALSE}
# PNI legacy series; the official page currently exposes data through 2022
pni_imunizacoes(uf = "MS")
pni_imunizacoes("cobertura", uf = "MS")

# Cervical and breast cancer exams
siscan(uf = "MS")
siscan("mamografia_residencia", uf = "MS", periodo = 2025)

# Historical SISVAN and financing tables
sisvan(uf = "MS")
financiamento_sus(uf = "MS")
```

Requests use HTTPS, an explicit response encoding, timeouts, and retries. The
defaults can be adjusted for slow connections:

```{r, eval = FALSE}
options(
  datasus.timeout = 30,
  datasus.download_timeout = 300,
  datasus.max_tries = 2,
  datasus.cache_ttl = 3600
)
```

## OpenDataSUS microdata

The package also discovers, downloads and reads modern surveillance
microdata from the official OpenDataSUS portal:

```{r, eval = FALSE}
# Search datasets and inspect their published resources
opendatasus_catalogo("dengue")
opendatasus_recursos("arboviroses-dengue")

# Read only a sample while exploring a large annual file
srag <- sivep_gripe(ano = 2025, n_max = 1000)
dengue <- sinan_dengue(ano = 2025, n_max = 1000)
cases <- sinan_mpox(ano = 2025, n_max = 1000)

# Contemporary e-SUS and PNI sources use their published partitions
adverse_events <- esavi(n_max = 1000)
mild_cases <- esus_sindrome_gripal(
  uf = "MS",
  ano = "last",
  n_max = 1000,
  colunas = c(
    "dataNotificacao", "municipioIBGE", "idade", "sexo"
  ),
  normalizar = TRUE
)
doses <- pni_doses(
  ano = "last",
  mes = "last",
  n_max = 1000,
  normalizar = TRUE
)
occupancy <- ocupacao_hospitalar(
  ano = "last",
  n_max = 1000,
  normalizar = TRUE
)

# Every result records its URL, retrieval time, local file and checksum
datasus_proveniencia(dengue)
datasus_validar_esquema(
  doses,
  "pni_doses",
  campos = c("data_vacinacao", "cnes")
)
```

Downloads are atomic and cached on disk. Set `atualizar = TRUE` to force a
fresh metadata query and download, or use `opendatasus_baixar()` when another
engine should read the file. `esus_sindrome_gripal()` is distinct from
hospitalized SRAG in `sivep_gripe()`, while `pni_doses()` provides
record-level files rather than the legacy aggregated TABNET series from
`pni_imunizacoes()`.

Some state resources are split into many physical files. Inspect their actual
parts with `opendatasus_arquivos()`. For files that should not be loaded in
memory, process bounded-size chunks:

```{r, eval = FALSE}
sg_resources <- opendatasus_recursos(
  "notificacoes-de-sindrome-gripal-leve-2020"
)
sg_ms <- sg_resources$id[
  sg_resources$formato == "CSV" &
    grepl("^Dados MS", sg_resources$nome)
]
opendatasus_arquivos(
  "notificacoes-de-sindrome-gripal-leve-2020",
  recurso = sg_ms,
  formato = "CSV"
)

totals <- opendatasus_processar(
  "notificacoes-de-sindrome-gripal-leve-2020",
  recurso = sg_ms,
  ano = NULL,
  colunas = c("municipioIBGE", "resultadoTeste"),
  tamanho_bloco = 50000,
  sistema = "sindrome_gripal",
  FUN = function(dados, posicao, arquivo) {
    table(dados$codigo_municipio_residencia)
  }
)
```

## Raw DBC/DBF microdata

Record-level SIM, SINASC and SIH files are available through a common API.
Start from the local catalog and inspect the official files before downloading:

```{r, eval = FALSE}
microdados_catalogo()
microdados_arquivos("sih", ano = 2024, mes = 1, uf = "AC")

admissions <- sih_microdados(
  ano = 2024,
  mes = 1,
  uf = "AC",
  colunas = c("MUNIC_RES", "DT_INTER", "DIAG_PRINC", "VAL_TOT"),
  n_max = 1000,
  normalizar = TRUE
)
```

The same workflow is exposed by `sim_microdados()` and
`sinasc_microdados()`. DBC files are decoded directly in memory; selecting
columns and limiting rows avoids allocating an entire large file during
exploration.

Common fields can be inspected and standardized conservatively:

```{r, eval = FALSE}
datasus_dicionario("sih")
datasus_padronizar(admissions, "sih")
datasus_proveniencia(admissions)
```

## Territorial reference

The package includes an offline IBGE hierarchy for current municipalities,
states and macroregions:

```{r, eval = FALSE}
# Discover current territories without a network request
datasus_territorios("regiao")
datasus_territorios("uf")
municipios_ms <- datasus_territorios("municipio", uf = "MS")

# Convert six-digit DATASUS codes to full seven-digit IBGE codes
normalizar_codigo_ibge(c("500270", "500370"))
extrair_codigo_ibge(c("500270 Campo Grande", "500370 Dourados"))

# Append municipality, state and region attributes
casos <- data.frame(
  codmun = c("500270", "500370"),
  ano = 2025,
  casos = c(10, 5)
)
adicionar_territorio(casos, "codmun")
```

Missing territory-period combinations can be created explicitly without
overwriting observed values:

```{r, eval = FALSE}
completar_territorios(
  casos,
  codigo = "codmun",
  periodo = "ano",
  periodos = 2023:2025,
  preencher = list(casos = 0)
)
```

The reference describes the current territorial hierarchy. Historical
observations are not automatically redistributed across boundary changes.

## Epidemiological analysis

Validated helpers cover frequent calculations after data retrieval:

```{r, eval = FALSE}
# Vectorized crude rates and exact Poisson confidence intervals
calcular_taxa(eventos = c(10, 25), populacao = c(10000, 20000))
intervalo_taxa(eventos = 10, populacao = 10000)

# Aggregate first, then calculate a grouped incidence rate
taxa_incidencia(
  dados,
  casos = "casos",
  populacao = "populacao",
  grupo = c("codigo_municipio", "ano"),
  confianca = 0.95
)

# Proportions, ratios and case fatality use the same validated engine
proporcao(dados, "vacinados", "elegiveis", grupo = "ano")
letalidade(dados, "obitos", "casos", grupo = "ano")

# Brazilian Sunday-to-Saturday epidemiological calendar
semana_epidemiologica(as.Date(c("2025-01-01", "2026-01-01")))
calendario_epidemiologico(2026)

# Right-aligned seven-observation moving average
media_movel(casos_diarios, janela = 7)
```

Population denominators can be joined without silently duplicating rows or
accepting ambiguous keys:

```{r, eval = FALSE}
dados <- juntar_populacao(
  eventos,
  denominadores,
  por = c(codmun = "codigo_municipio", ano = "ano"),
  coluna_populacao = "habitantes"
)
taxa_mortalidade(dados, "obitos", "populacao", grupo = c("codmun", "ano"))
```

Direct age standardization accepts named standard-population weights and can
calculate separate results for years, municipalities or other groups. WHO,
Segi and Scandinavian standards are ready to use:

```{r, eval = FALSE}
padronizar_idade(
  eventos = obitos_por_idade,
  populacao = habitantes_por_idade,
  idade = faixa_etaria,
  populacao_padrao = populacao_padrao("oms"),
  grupo = ano,
  confianca = 0.95
)
```

## Vignettes

Five guides provide complete learning paths through the package:

```{r, eval = FALSE}
vignette("Introduction_to_datasus", package = "datasus")
vignette("accessing-datasus", package = "datasus")
vignette("modern-surveillance", package = "datasus")
vignette("geography-and-analysis", package = "datasus")
vignette("large-files-and-microdata", package = "datasus")
```

The four task-oriented guides cover TABNET access, modern OpenDataSUS
surveillance, territorial and epidemiological analysis, and efficient
processing of large microdata files.
