---
title: "Accessing DATASUS tables"
author: "Renato Prado Siqueira"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Accessing DATASUS tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
library(datasus)
```

## Overview

`datasus` provides catalog-driven interfaces to the tables published through
TABNET. The main workflow is:

1. discover the system and dataset with `datasus_catalogo()`;
2. inspect the dimensions and filters with `datasus_opcoes()`;
3. submit a focused query through the corresponding function;
4. retain the provenance attached to the result.

Network examples are not evaluated when this vignette is built because the
availability and response time of DATASUS are outside the package's control.

## Discover datasets

The local catalog can be searched without contacting DATASUS:

```{r catalog, eval=FALSE}
datasus_catalogo()
datasus_catalogo("mortalidade")
datasus_catalogo("cnes")
datasus_catalogo("sinan")
```

The result identifies the system, dataset and public function to use. Once a
dataset has been selected, inspect the current TABNET form:

```{r options, eval=FALSE}
options <- datasus_opcoes(
  sistema = "sim",
  conjunto = "obitos",
  abrangencia = "uf"
)

options$linha
options$coluna
options$conteudo
names(options$filtros)
```

Using the labels returned by `datasus_opcoes()` avoids embedding assumptions
about a form that the portal may change later.

## Vital statistics

`sim()` retrieves mortality data and `sinasc()` retrieves live-birth data.
Both accept a year, `"last"` for the latest available period, and geographic
or demographic filters.

```{r vital-statistics, eval=FALSE}
deaths <- sim(
  conjunto = "obitos",
  abrangencia = "uf",
  periodo = 2024,
  coluna = "Ano do óbito"
)

male_deaths <- sim(
  conjunto = "obitos",
  uf = "MS",
  periodo = 2024,
  filtros = list(sexo = "Masculino")
)

births <- sinasc(
  uf = "MS",
  periodo = 2024,
  coluna = "Ano do nascimento"
)
```
The older `sim_*()` and `sinasc_*()` functions remain as compatibility
wrappers, but new code should use the unified functions above.

## Health services and population

Hospital, ambulatory and establishment tables follow the same conventions:

```{r health-services, eval=FALSE}
admissions <- sih_producao(
  uf = "MS",
  conteudo = "Internações",
  periodo = 2025,
  filtros = list(carater_atendimento = "Urgência")
)

procedures <- sia_producao(
  uf = "MS",
  conteudo = "Qtd.aprovada",
  periodo = 2025
)

beds <- cnes(
  conjunto = "leitos_internacao",
  uf = "MS",
  periodo = "last"
)

population <- populacao_residente(
  uf = "MS",
  periodo = 2021
)
```

Use `sih_morbidade()` when the analysis is diagnosis-oriented rather than
production-oriented:

```{r morbidity, eval=FALSE}
morbidity <- sih_morbidade(
  uf = "MS",
  linha = "Capítulo CID-10",
  conteudo = "Internações",
  periodo = 2025
)
```

## Surveillance, screening and financing

The catalog also covers disease-specific SINAN tables, immunization,
nutritional surveillance, cancer screening and SUS financing:

```{r other-tabnet, eval=FALSE}
dengue <- sinan("dengue", uf = "MS", periodo = 2025)

coverage <- pni_imunizacoes(
  conjunto = "cobertura",
  uf = "MS"
)

mammograms <- siscan(
  conjunto = "mamografia_residencia",
  uf = "MS",
  periodo = 2025
)

nutrition <- sisvan(uf = "MS")
financing <- financiamento_sus(uf = "MS")
```

## Provenance and reproducibility

Results carry a provenance record with the requested system, filters, source
URL and retrieval time:

```{r tabnet-provenance, eval=FALSE}
source <- datasus_proveniencia(deaths)
str(source)
```

For a reproducible analysis, save the query arguments together with the
result, request explicit years instead of `"last"`, and record the package
version:

```{r reproducibility, eval=FALSE}
analysis_metadata <- list(
  package_version = as.character(packageVersion("datasus")),
  query = list(
    sistema = "sim",
    conjunto = "obitos",
    abrangencia = "uf",
    periodo = 2024
  ),
  provenance = datasus_proveniencia(deaths)
)
```
